Supplementary Materials

### 2023-06-29 manuscript/supplementary.tex
methods

\section{Quantitative Genetic Linked Selection Model Theory}

@@

\subsection{Heritable Fitness and the Accumulation of Autocovariance}

@@

\begin{align}
  \label{supp-eq:b}
  B(x, \Psi) &= \prod_g^S \exp \left( \int_0^1 b(\mu(s, k(g)), s, L_g, r_g, d(x, g)) ds \right) \nonumber  \\
             &= \exp \left( \sum_g^S \int_0^1 b(\mu(s, k(g)), s, L_g, r_g, d(x, g)) ds \right)
\end{align}
%
where, $b(\mu(s, k(g)), s, L_g, r_g, d(x,g))$ is the log reduction due to
segment $g$, and there are $S$ total segments. Here, $\mu(s, k(g))$ is the
per-basepair per-genome rate that mutations with selection coefficient $s$
enter segments with feature class $k(g)$ (e.g. it incorporates the distribution
of fitness effects and mutation rate), $L_g$ and $r_g$ are the length in
basepairs and recombination rate per basepair of segment $g$, and the
recombination distance between focal site $x$ and segment $g$ is $d(x, g)$. The
functional form of $b$ varies depending on whether the classic BGS model is
used (Supplementary Section \ref{supp:reduction-classic}), or our form of
Santiago and Caballero's (\citeyear{Santiago2016-mu}) equations (see
Supplementary Section \ref{supp:reduction-sc16}).

@@

\begin{align}
  \label{supp-eq:b-disc}
  B(x, \Psi) = \exp \left( \sum_g^S \sum_{i}^{m_s} b(\mu_{i, \;k(g)}, s_i, L_g, r_g, d(x, g))\right).
\end{align}
%
Because it is computationally-intensive to numerically solve the system of two
non-linear equations to calculate $b$ at each segment, these are pre-computed
for an $m_s$-element grid of selection coefficients and an $m_\mu$-element grid
of mutation rates--DFE products. Likewise, when we calculate $b$ under classic
BGS models, we pre-calculate segment components across these same grids, and
combine them for each $x$.

@@

\begin{align}
  \mathbf{M}_\text{S} = \mu \begin{bmatrix}
    w_{10^{-6},\text{CDS}} & w_{10^{-6},\text{UTR}} & w_{10^{-6},\text{PC}} \\
    w_{10^{-5},\text{CDS}} & w_{10^{-5},\text{UTR}} & w_{10^{-5},\text{PC}} \\
    \vdots & \vdots & \vdots \\
    w_{10^{-1},\text{CDS}} & w_{10^{-1},\text{UTR}} & w_{10^{-1},\text{PC}} \\
  \end{bmatrix}
\end{align}
%
which has $2 + K \times (m_s-1)$ free parameters, since each column must sum to
one. This imposes the constraint that $\sum_i w_{i,k} = 1$, requiring
constrained MLE optimization. We implement this using softmax, which describe 
further in Supplementary Section \ref{supp:optim}.

@@

Note that the asymptotic model above determines the reduction in effective
population size experienced by a focal site in the middle of a $L$-basepair
segment, but B maps require the reduction $b'$ experienced by a focal site $x$
at an arbitrary recombination distance apart. For each segment, we pre-compute
the solution to the system of two equations in Supplementary Equation
\eqref{supp-eq:sc16-sys} for the mutation and selection coefficient grids,
giving us $\widetilde{V}$ and $\widetilde{Q}_\infty^2$ for each set of
parameters. Then,

\begin{align} 
    b'(\mu, s, L, r, d) = -\frac{\widetilde{V}}{2} \left(\frac{1}{(1-(1-d)Z}\right)^2 
\end{align}
%
where $Z = 1 - Us / (U-\widetilde{R})$ and $\widetilde{V} = Us - 2\widetilde{R}s$. As
before, $d$ is the recombination fraction between the segment and the focal
neutral site.

@@

\begin{align}
  \pi_l = \frac{n_D}{n_T}
\end{align}

which is identical to the more common expression of this estimator,

\begin{align}
  \pi_l = \frac{2}{n(n-1)} \sum_{i < j}^n k_{i,j}
\end{align}

where $k_{i,j}$ is 1 if the alleles at this site differ at site $l$, and 0
otherwise.

@@

\begin{align}
  \label{eq:}
  \pi^{(1)} = \frac{1}{L} \sum_{i=1}^L \frac{n_{D,i}}{n_{T,i}}
\end{align}

which if the number of samples across loci is constant, simplifies to an
unweighted average across sites. Second, one can take a weighted average, with
weights determined by the total number of samples present at a site,

\begin{align}
  \label{eq:}
  \pi^{(2)} = \frac{1}{\sum_{i=1}^L n_i} \sum_{i=1}^L n_i \frac{n_{D,i}}{n_{T,i}}
\end{align}

Third, one can weight by the number of pairwise comparisons at a site,
$n_{T,i}$, rather than total number of samples, $n_i$, which leads to a ratio
of sums,

@@

We predominantly use the estimator $\pi^{(3)}$, as it corresponds to how we
summarize the matrix $\mathbf{Y}$ across windows for our likelihood. All
methods have mean squared errors and biases very close to one another (TODO).

@@

  \caption{The distribution of diversity across the genome at the megabase
  scale, with outliers trimmed. The blue line is the normal CDF, fit with MLE
parameters XXX. The inset figure is the untrimmed data, with trimmed points
shown in red.}

  \label{suppfig:trimming}
\end{figure}

@@

The rescaled $B'$ calculation outputs a $m_\mu \times m_s \times S$
multidimensional array $\mathbf{R}$ of ratchet rates per segment. These are
rescaled by the segment lengths $\mathbf{l}$, giving the per-basepair mutation
rate array. The per-basepair ratchet rate given maximum likelihood estimates of
$\hat{\mu}$ and $\widehat{\mathbf{W}}$ is used to predicted the ratchet rate
for each segment $g$. The DFE estimate for feature $k$ is a column vector of
$\widehat{\mathbf{W}}$, i.e. $\widehat{\mathbf{w}}_k$

@@

>>>>

\section{Quantitative Genetic Linked Selection Model Theory}
\label{suppsec:theory}

@@

\subsection{Heritable Fitness and the Accumulation of Autocovariance}
\label{suppsec:heritable-fitness}

@@

\begin{align}
  \label{supp-eq:b}
  B(x, \Psi) &= \prod_{g \in G} \exp \left( \int_0^1 b(\mu(s, k(g)), s, L_g, r_g, d(x, g)) ds \right) \nonumber  \\
             &= \exp \left( \sum_{g \in G} \int_0^1 b(\mu(s, k(g)), s, L_g, r_g, d(x, g)) ds \right)
\end{align}
%
where, $b(\mu(s, k(g)), s, L_g, r_g, d(x,g))$ is the log reduction due to
segment $g$, and there are $S$ total segments. Here, $\mu(s, k(g))$ is the
per-basepair per-genome rate that mutations with selection coefficient $s$
enter segments with feature class $k(g)$ (e.g. it incorporates the distribution
of fitness effects and mutation rate), $L_g$ and $r_g$ are the length in
basepairs and recombination rate per basepair of segment $g$, and the
recombination distance between focal site $x$ and segment $g$ is $d(x, g)$. The
functional form of $b$ varies depending on whether the classic BGS model is
used (Supplementary Section \ref{supp:reduction-classic}), or our form of
Santiago and Caballero's (\citeyear{Santiago2016-mu}) equations (see
Supplementary Section \ref{supp:reduction-sc16}).

@@

\begin{align}
  \label{supp-eq:b-disc}
  B(x, \Psi) = \exp \left( \sum_g^S \sum_{i}^{m_s} b(\mu_{i, \;k(g)}, s_i, L_g, r_g, d(x, g))\right).
\end{align}
%
Because it is computationally-intensive to numerically solve the system of two
non-linear equations to calculate $b$ at each segment, these are pre-computed
for an $n_s$-element grid of selection coefficients and an $n_\mu$-element grid
of mutation rates--DFE products. Likewise, when we calculate $b$ under classic
BGS models, we pre-calculate segment components across these same grids, and
combine them for each $x$.

@@

\begin{align}
  \mathbf{M}_\text{S} = \mu \begin{bmatrix}
    w_{10^{-6},\text{CDS}} & w_{10^{-6},\text{UTR}} & w_{10^{-6},\text{PC}} \\
    w_{10^{-5},\text{CDS}} & w_{10^{-5},\text{UTR}} & w_{10^{-5},\text{PC}} \\
    \vdots & \vdots & \vdots \\
    w_{10^{-1},\text{CDS}} & w_{10^{-1},\text{UTR}} & w_{10^{-1},\text{PC}} \\
  \end{bmatrix}
\end{align}
%
which has $2 + K \times (n_s-1)$ free parameters, since each column must sum to
one. This imposes the constraint that $\sum_i w_{i,k} = 1$, requiring
constrained MLE optimization. We implement this using softmax, which describe 
further in Supplementary Section \ref{supp:optim}.

@@

Note that the asymptotic model above determines the reduction in effective
population size experienced by a focal site in the middle of a $L$-basepair
segment, but B maps require the reduction $b'$ experienced by a focal site $x$
at an arbitrary recombination distance apart. For each segment, we pre-compute
the solution to the system of two equations in Supplementary Equation
\eqref{supp-eq:sc16-sys} for the mutation and selection coefficient grids,
giving us $\widetilde{V}_a$ and $\widetilde{Q}_\infty^2$ for each set of
parameters. Then,

\begin{align} 
    b'(\mu, s, L, r, d) &= -\frac{\widetilde{V}_a(\mu, s, r)}{2} Q^2(\mu, s, r) \\
                        Q^2(\mu, s, r)&= \left(\frac{1}{(1-(1-d)\widetilde{Z}}\right)^2 
\end{align}
%
where $\widetilde{Z} = 1 - Us / (U-\widetilde{R})$ and $\widetilde{V}_a = Us - 2\widetilde{R}s$. As
before, $d$ is the recombination fraction between the segment and the focal
neutral site.

@@

\begin{align}
  \pi_l = \frac{n_D}{n_T}
\end{align}
%
which is identical to the more common expression of this estimator,

\begin{align}
  \pi_l = \frac{2}{n(n-1)} \sum_{i < j}^n k_{i,j}
\end{align}
%
where $k_{i,j}$ is 1 if the alleles at this site differ at site $l$, and 0
otherwise.

@@

\begin{align}
  \label{eq:}
  \pi^{(1)} = \frac{1}{L} \sum_{i=1}^L \frac{n_{D,i}}{n_{T,i}}
\end{align}
%
which if the number of samples across loci is constant, simplifies to an
unweighted average across sites. Second, one can take a weighted average, with
weights determined by the total number of samples present at a site,

\begin{align}
  \label{eq:}
  \pi^{(2)} = \frac{1}{\sum_{i=1}^L n_i} \sum_{i=1}^L n_i \frac{n_{D,i}}{n_{T,i}}
\end{align}
%
Third, one can weight by the number of pairwise comparisons at a site,
$n_{T,i}$, rather than total number of samples, $n_i$, which leads to a ratio
of sums,

@@

We predominantly use the estimator $\pi^{(3)}$, as it corresponds to how we
summarize the matrix $\mathbf{Y}$ across windows for our likelihood. All
methods have mean squared errors and biases very close to one another.

@@

@@

The rescaled $B'$ calculation outputs a $n_\mu \times n_s \times S$
multidimensional array $\mathbf{R}$ of ratchet rates per segment. These are
rescaled by the segment lengths $\mathbf{l}$, giving the per-basepair mutation
rate array. The per-basepair ratchet rate given maximum likelihood estimates of
$\hat{\mu}$ and $\widehat{\mathbf{W}}$ is used to predicted the ratchet rate
for each segment $g$. The DFE estimate for feature $k$ is a column vector of
$\widehat{\mathbf{W}}$, i.e. $\widehat{\mathbf{w}}_k$

@@

  \caption{The distribution of diversity across the genome at the megabase
  scale, with outliers trimmed. The blue line is the normal CDF, fit with MLE
parameters XXX. The inset figure is the untrimmed data, with trimmed points
shown in red.}
  \label{suppfig:trimming}
\end{figure}
